Examining spatial equity in accessibility to a public bicycle share program using a balanced floating catchment area approach

### 2021-03-28 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd

---
title: Examining spatial equity in accessibility to a public bicycle share program using a balanced floating catchment area approach 
author:
  - name: Elise Desjardins
    email: desjae@mcmaster.ca
    affiliation: Some School
    footnote: 1
  - name: Christopher D. Higgins
    email: cd.higgins@utoronto.ca
    affiliation: Another University
    footnote: 2
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: Some School
    footnote: 2
address:
  - code: McMaster University
    address: School of Earth, Environment & Society, 1280 Main Street West, Hamilton, ON L8S4L8
  - code: University of Toronto Scarborough
    address: Department of Geography & Planning, 1265 Military Trail, Toronto, ON M1C1A4
footnote:
  - code: 1
    text: "Corresponding Author"
  - code: 2
    text: "Equal contribution"
abstract:

  Public bicycle share programs (PBSPs) are implemented to promote cycling as a convenient and sustainable mode of transportation. These systems can also play a role in addressing transportation needs and advancing transportation equity if they make cycling more accessible to lower income or disadvantaged populations. In Ontario, the City of Hamilton launched a public bicycle share program in 2015 that currently has over 900 operational bicycles and 130 docking stations. This system was expanded in 2018 by an equity initiative that added twelve docking stations with the explicit objective of increasing spatial equity in access. Since the cost of reaching a bicycle share station decreases the potential of accessing the program, the location of stations matters, and previous research found that Hamilton’s public bicycle share program targeted well disadvantaged areas in the city. In this research, we revisit the case of Hamilton and investigate differentials in accessibility to bicycle share stations using a balanced floating catchment area (BFCA) accessibility approach. We compare accessibility  with and without the equity stations to assess the effect of the initiative. We implement our analysis parting from micro zones to better reflect walking to a docking station, and conduct a sensitivity analysis at several walking time thresholds. We then reaggregate the estimated accessibility for further analysis using census data. The addition of equity stations increased the serviced population at every threshold examined, and although accessibility increased for the whole population, this increase was relatively modest especially for population in the bottom 20% of median household income.

@@

```{r load-packages, include=FALSE}
library(biscale)
library(cowplot)
library(data.table)
library(dplyr)
library(gdistance)
library(gridExtra)
library(kableExtra)
library(patchwork)
library(readr)
library(raster)
library(rgdal)
library(sf)
library(sobiEquity)
library(tidyverse)
library(tinytex)
library(units)
library(disk.frame)
library(pycno)
library(r5r) # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
```

```{r generate-package-bibliography, include=FALSE}
knitr::write_bib(.packages(), "packages.bib")
```

@@

The potential of public bicycle share programs (PBSPs) to increase cycling levels is but one of many reasons for implementing such programs in urban areas [@hosfordEvaluationImpactPublic2018; @hosfordEvaluatingImpactImplementing2019]. As a healthy, inexpensive, and convenient form of public transportation, shared bicycles can encourage individuals to take up cycling for short local trips or first and last mile trips to other public transportation instead of using personal vehicles. These systems can also play a role in addressing transportation needs and advancing transportation equity if they make cycling more accessible to lower income or disadvantaged populations. As these systems become increasingly common, their introduction has been accompanied by a flurry of research that investigates them from the perspective of spatial and transportation equity [e.g., @smith2015exploring; @hosfordWhoArePublic2018; @mooneyFreedomStationSpatial2019; @hullgrassoBikeShareEquity2020; @qianBikesharingEquityDisadvantaged2020; @qianEnhancingEquitableService2020]. Indeed, although these programs are available to the general public and ought to be accessible to any individuals who wish to use them, research on PBSPs located in North American and European cities indicates that there are inequities in spatial accessibility in need of redress.

In this paper we investigate the case of the PBSP in the City of Hamilton, in Ontario, Canada. Hamilton Bike Share (previously also known as Social Bicycles, or SoBi) was launched in 2015 and currently has over 900 operational bicycles and 130 docking stations. An interesting feature of this system is that it was expanded by an equity initiative that introduced twelve docking stations in 2018 with the explicit objective of increasing spatial accessibility. Hamilton Bike Share was already studied by @hosfordWhoArePublic2018 as part of a selection of PBSPs in Canada. These authors found that most of the cities in their research could benefit of greater efforts to expand service to areas with lower socio-economic populations - but Hamilton fared somewhat better than other cities in this respect. What distinguishes our research is the use of a balanced floating catchment area (BFCA) accessibility approach that accounts for the supply of stations and the potential demand from the population serviced. This requires a more disaggregated approach than the use of census geographies [q.v., @hosfordWhoArePublic2018], since many trips to a bike share station likely happen at a level lower than even the smallest census geography. For this reason, we implement our analysis parting from micro population zones to better reflect the friction of walking to a docking station. Further, we conduct a sensitivity analysis at several walking time thresholds. In terms of spatial equity, we compare accessibility with and without the equity stations to assess the effect of the initiative, before reaggregating the data for further analysis using median household income information from the census. We find that the addition of equity stations does increase the serviced population at every threshold examined. However, although accessibility increased for the whole population, this increase was relatively modest, especially for population in the bottom 20% of median household income. The analysis also identifies areas under-serviced within the core service area of the system.

@@

Accessibility has been applied in both a positive and normative way to inform transportation planning [@paezMeasuringAccessibilityPositive2012], but its utility to this field has evolved over the past century and has increasingly become linked with recent interest in prioritizing local proximity and modes that are suitable for local trips like walking and cycling [@levineCenturyEvolutionAccessibility2020]. As a measure of the ease of reaching potential destinations spread spatially in a given area, accessibility is relevant to PBSPs because it can identify current inequities in the provision of infrastructure, as well as guide interventions that increase access for groups that are under-serviced or address gaps in transportation options. It also addresses some of the challenges of other performance measures such as level of service within a transportation network by measuring person-based indicators and exploring differences in use between population groups [@paezMeasuringAccessibilityPositive2012].

Beyond the utility derived from using shared bikes to destinations of value, an important aspect of accessibility associated with PBSPs is the distance an individual must travel to reach a bicycle share station [@kabraBikeShareSystemsAccessibility2020; @wangNewBikeShare2019]. Since the time or distance needed to reach a bicycle share station decreases the potential of accessing the program and ultimately their use of the program, the location and size (e.g., maximum number of bicycle racks available) of stations matters. Indeed, distance to bicycle share stations is associated with use [@fullerUseNewPublic2011; @wangNewBikeShare2019] and can be a barrier to using PBSPs [@fishmanBarriersBikesharingAnalysis2014]. Kabra et al. [-@kabraBikeShareSystemsAccessibility2020] found that the majority of bike share usage in Paris comes from areas within 300 metres of stations, which amounts to 2-4 minutes walking by an able-bodied adult. Furthermore, similar to other public amenities affected by crowding, like health care services [e.g., @pereiraGeographicAccessCOVID192021], the utility of stations is limited by the number of bicycles that they can hold. The program may not necessarily be improved if stations are easy to reach but offer only a small number of bicycles. Likewise, more people may not opt to use the program if the supply of bicycles available at the nearest station is insufficient to meet demand. The location and size of stations is important to increase the utility of this public transportation option for more people, thus achieving vertical equity. Accessibility analyses for PBSPs constitute a positive and evaluation-based approach that also has the potential to inform equity efforts. For instance, Wang and Lindsey [-@wangNewBikeShare2019] investigated whether new or relocated bicycle share stations increased accessibility and use, which offered important insights to improve the performance of the program.

@@

In contrast, floating catchment area (FCA) methods incorporate information on capacity and demand in calculating accessibility. FCA measures have been widely employed in healthcare accessibility research and are typically calculated across two steps. In the first, a ratio of supply to demand at service locations is calculated, such as the number of beds at a hospital divided by the number of people within the catchment area of the hospital, weighted by the distance involved in reaching the facility. Next, these service level ratios are allocated back to the population centres and summarized as a measure of congested accessibility. Thus, this model does a good job of considering potential crowding or competition for services. While there have been many methodological innovations in FCA methods [for example, @delamaterSpatialAccessibilitySuboptimally2013; @luoMeasuresSpatialAccessibility2003; @luoEnhancedTwostepFloating2009; @radkeSpatialDecompositionsModeling2000; @wanThreestepFloatingCatchment2012], a recent improvement to this approach was achieved through a simple and intuitive balancing of the impedance that addressed the effects of demand and service inflation found in previous FCA approaches [see @paezDemandLevelService2019].

When measuring accessibility, researchers have taken different approaches when it comes to aggregation of data, either by using the individual or household as the smallest unit of analysis or larger spatial zones. Previous research on bike share equity has typically used a meso- or macro-level approach with aggregate data from entire neighbourhoods or census tracts [@babagoliExploringHealthSpatial2019; @mooneyFreedomStationSpatial2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], although there are recent exceptions [@chenExploringEquityPerformance2019; @chenUnobservedHeterogeneityTransportation2021]. This is also true for studies examining correlates of bike share demand [@wangNewBikeShare2019]. Handy and Niemeier [-@handyMeasuringAccessibilityExploration1997] note that using disaggregated data in accessibility analyses provides a more accurate estimate for individuals, which is useful for addressing vertical inequities in PBSP usage.

Using the balanced floating catchment area method [@paezDemandLevelService2019], a novel approach that has not been used yet in cycling research, with disaggregate population-level data, in this paper we examine accessibility to a public bicycle share program in Hamilton, Ontario. We (1) conduct a sensitivity analysis by measuring accessibility and level of service to bike share stations at different walking time thresholds to reach a station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes; (2) explore the contribution of specific stations that were added to Hamilton's PBSP to reducing both horizontal and vertical inequities; and (3) examine whether disparities in accessibility exist according to median household income of dissemination areas within the core service area.

@@

In fact, since the proportional allocation procedure means that any proportion of the population allocated to a station is never allocated to other stations, and conversely any level of service allocated to a population is never re-allocated elsewhere, this property is replicated for any level of aggregation. For this paper, we employ a hybrid location-based and person-based approach to calculating accessibility using disaggregate population data. In their review of accessibility measures, Geurs and van Wee [-@geursAccessibilityEvaluationLanduse2004] highlight the need for greater inclusion of individual spatio-temporal constraints but acknowledge the challenges of acquiring and analyzing person-based data. This comes after Kwan's [-@kwanSpaceTimeIntegral1998] work to show that space-time measures are  better able of capturing interpersonal differences, especially the effect of space-time constraints on individual behaviour, and are more helpful for unraveling gender/ethnic differences. Applying the balanced floating catchment area approach allows us to examine accessibility by stratifying according to median household income, which would constitute the individual component of the accessibility measure [@geursAccessibilityEvaluationLanduse2004]. However, conducting a further sensitivity analysis to measure accessibility at different distance thresholds would help to consider potential spatio-temporal constraints. Different people may be willing to travel different distances to access a public bicycle share station.

@@

```{r interpolated-population, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Interpolated population in the Hamilton Bike Share core service area (outlined in black) and within 30 minutes of walking to the core service area (outlined in blue). Each micro population cell is 50-by-50 m in size. The downtown area is outlined in dark green."}
ggplot() + 
  geom_sf(data = cropped_da,
          fill = "gray",
          color = NA) + 
  geom_sf(data = population_50x50,
          aes(color = population)) +
  scale_color_distiller(palette = "YlOrRd", 
                        direction = 1) +
  geom_sf(data = sobi_service,
          fill = NA,
          colour = "black") +
  geom_sf(data = community_downtown %>%
            dplyr::filter(NAME == "Hamilton"),
          fill = NA,
          color = "dark green") +
  theme(panel.background = element_rect(fill = "white"))
```

@@

 ```{r figure-6, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 3 minutes walk (minimum threshold) compared between current system with equity stations and the original system without equity stations."}
ggplot() +
  geom_sf(data = population_50x50,
          color = "gray") +
  geom_sf(data = accessibility %>%
            dplyr::filter(threshold == "3 min"),
          aes(color = accessibility)) +
  geom_sf(data = sobi_service,
          fill = NA) +
  scale_color_distiller(palette = "PuRd", 
                        direction = 1) +
  facet_grid(system ~ threshold)  +
  geom_sf(data = community_downtown %>%
            dplyr::filter(NAME == "Hamilton"),
          fill = NA,
          color = "dark green") +
  # geom_sf(data =  sobi_hubs %>% 
  #           dplyr::filter(hub_type == "ERI"), 
  #         size = 1.5, shape = 21, 
  #         fill = "green") +
  theme(panel.background = element_rect(fill = "white"))
```

@@

```{r figure-7, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 5 minutes walk (average threshold) compared between current system with equity stations and the original system without equity stations."}
#accessibility plot for current system with equity stations at 5 minutes walk
ggplot() +
  geom_sf(data = population_50x50,
          color = "gray") +
  geom_sf(data = accessibility %>%
            dplyr::filter(threshold == "5 min"),
          aes(color = accessibility)) +
  geom_sf(data = sobi_service,
          fill = NA) +
  scale_color_distiller(palette = "PuRd", 
                        direction = 1) +
  facet_grid(system ~ threshold)  +
  geom_sf(data = community_downtown %>%
            dplyr::filter(NAME == "Hamilton"),
          fill = NA,
          color = "dark green") +
  # geom_sf(data =  sobi_hubs %>% 
  #           dplyr::filter(hub_type == "ERI"), 
  #         size = 1.5, shape = 21, 
  #         fill = "green") +
  theme(panel.background = element_rect(fill = "white"))
```

@@

```{r figure-8, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 10 minutes walk (maximum threshold) compared between current system with equity stations and the original system without equity stations."}
#accessibility plot for current system with equity stations at 10 minutes walk
ggplot() +
  geom_sf(data = population_50x50,
          color = "lightgray") +
  geom_sf(data = accessibility %>%
            dplyr::filter(threshold == "10 min"),
          aes(color = accessibility)) +
  geom_sf(data = sobi_service,
          fill = NA) +
  scale_color_distiller(palette = "PuRd", 
                        direction = 1) +
  facet_grid(system ~ threshold)  +
  geom_sf(data = community_downtown %>%
            dplyr::filter(NAME == "Hamilton"),
          fill = NA,
          color = "dark green") +
  # geom_sf(data =  sobi_hubs %>% 
  #           dplyr::filter(hub_type == "ERI"), 
  #         size = 1.5, shape = 21, 
  #         fill = "green") +
  theme(panel.background = element_rect(fill = "white"))
```

@@

```{r figure-9, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 15 minutes walk (extreme threshold) compared between current system with equity stations and the original system without equity stations."}
#accessibility plot for current system with equity stations at 15 minutes walk
ggplot() +
  geom_sf(data = population_50x50,
          color = "lightgray") +
  geom_sf(data = accessibility %>%
            dplyr::filter(threshold == "15 min"),
          aes(color = accessibility)) +
  geom_sf(data = sobi_service,
          fill = NA) +
  scale_color_distiller(palette = "PuRd", 
                        direction = 1) +
  facet_grid(system ~ threshold)  +
  geom_sf(data = community_downtown %>%
            dplyr::filter(NAME == "Hamilton"),
          fill = NA,
          color = "dark green") +
  # geom_sf(data =  sobi_hubs %>% 
  #           dplyr::filter(hub_type == "ERI"), 
  #         size = 1.5, shape = 21, 
  #         fill = "green") +
  theme(panel.background = element_rect(fill = "white"))
```

## Accessibility by Median Household Income

One of the unique properties of the balanced floating catchment area method is that data can be reaggregated while preserving the total population in the service area and the supply at each station. This avoids demand and supply inflation, and also enables us to present findings in a way that is perhaps more intuitive to interpret. Therefore, we reaggregate population and accessibility from the 50-by-50 m population cells to the dissemination area. We then can compare accessibility to median household income statistics from the census.

Panels 1 [\ref{fig:figure-bi-map-threshold-3}], 2 [\ref{fig:figure-bi-map-threshold-5}], 3 [\ref{fig:figure-bi-map-threshold-10}], and 4 [\ref{fig:figure-bi-map-threshold-15}] depict bivariate choropleth maps with the combine the spatial distribution of accessibility and median household income, using tertiles for the coloring scheme. As expected, the analysis indicates that the extreme threshold of fifteen minutes is associated with the largest number of people who are within the assumed service areas of stations. We find that stations added to Hamilton's public bicycle share program to increase equity for disadvantaged neighbourhoods achieved their goal by increasing access

>>>>

---
title: Examining spatial equity in accessibility to a public bicycle share program using a balanced floating catchment area approach 
author:
  - name: Elise Desjardins
    email: desjae@mcmaster.ca
    affiliation: McMaster University
    footnote: 1
  - name: Christopher D. Higgins
    email: cd.higgins@utoronto.ca
    affiliation: University of Toronto Scarborough
    footnote: 
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    footnote: 
address:
  - code: McMaster University
    address: School of Earth, Environment & Society, 1280 Main Street West, Hamilton, ON L8S4L8
  - code: University of Toronto Scarborough
    address: Department of Geography & Planning, 1265 Military Trail, Toronto, ON M1C1A4
footnote:
  - code: 1
    text: "Corresponding Author"
abstract:

  Public bicycle share programs (PBSPs) are implemented to promote cycling as a convenient and sustainable mode of transportation. These systems can also play a role in addressing transportation needs and advancing transportation equity if they make cycling more accessible to lower income or disadvantaged populations. In Ontario, the City of Hamilton launched a public bicycle share program in 2015 that currently has over 900 operational bicycles and 130 docking stations. This system was expanded in 2018 by an equity initiative that added twelve docking stations with the explicit objective of increasing spatial equity in access. Since the cost of reaching a bicycle share station decreases the potential of accessing the program, the location of stations matters, and previous research found that Hamilton’s public bicycle share program targeted well disadvantaged areas in the city. In this research, we revisit the case of Hamilton and investigate differentials in accessibility to bicycle share stations using a balanced floating catchment area (BFCA) accessibility approach. We compare accessibility  with and without the equity stations to assess the effect of the initiative. We implement our analysis parting from micro zones to better reflect walking to a docking station, and conduct a sensitivity analysis at several walking time thresholds. We then reaggregate the estimated accessibility for further analysis using census data. Our findings indicate that the addition of equity stations increased the horizontal equity of the system by growing the serviced population at every threshold examined. With respect to vertical equity, although accessibility increased for the whole population, this increase was relatively modest especially for population in the bottom 20% of median total household income.

@@

```{r load-packages, include=FALSE}
library(biscale)
library(cowplot)
library(data.table)
library(disk.frame)
library(dplyr)
library(gdistance)
library(gridExtra)
library(kableExtra)
library(pycno)
library(r5r) # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.htmllibrary(readr)
library(raster)
library(rgdal)
library(sf)
library(sobiEquity)
library(tidyverse)
library(tinytex)
library(units)
```

```{r generate-package-bibliography, include=FALSE}
knitr::write_bib(c(.packages(), "knitr", "rticles"), "packages.bib")
```

@@

The potential of public bicycle share programs (PBSPs) to increase cycling levels is but one of many reasons for implementing such programs in urban areas [@hosfordEvaluationImpactPublic2018; @hosfordEvaluatingImpactImplementing2019]. As a healthy, inexpensive, and convenient form of public transportation, shared bicycles can encourage individuals to take up cycling for short local trips or first and last mile trips to other public transportation instead of using personal vehicles. These systems can also play a role in addressing transportation needs and advancing transportation equity if they make cycling more accessible to lower income or disadvantaged populations. As these systems become increasingly common, their introduction has been accompanied by a flurry of research that investigates them from the perspective of spatial and transportation equity [e.g., @smith2015exploring; @hosfordWhoArePublic2018; @mooneyFreedomStationSpatial2019; @hullgrassoBikeShareEquity2020; @qianBikesharingEquityDisadvantaged2020; @qianEnhancingEquitableService2020]. Indeed, although these programs are available to the general public and ought to be accessible to any individuals who wish to use them, research on PBSPs located in North American and European cities indicates that inequity in spatial accessibility needs redressing.

In this paper we investigate the case of the PBSP in the City of Hamilton, in Ontario, Canada. Hamilton Bike Share (previously also known as Social Bicycles, or SoBi) was launched in 2015 and currently has over 900 operational bicycles and 130 docking stations. An interesting feature of this system is that it was expanded by an equity initiative that introduced twelve docking stations in 2018 with the explicit objective of increasing spatial accessibility. Hamilton Bike Share was already studied by @hosfordWhoArePublic2018 as part of a selection of PBSPs in Canada. These authors found that most of the cities in their research could benefit of greater efforts to expand service to areas with lower socio-economic populations - but Hamilton fared somewhat better than other cities in this respect. What distinguishes our research is the use of a balanced floating catchment area (BFCA) accessibility approach that accounts for the supply of stations and the potential demand from the population serviced. This requires a more disaggregated approach than the use of census geographies [q.v., @hosfordWhoArePublic2018], since many trips to a bike share station likely happen at a level lower than even the smallest census geography. For this reason, we implement our analysis parting from micro population zones to better reflect the friction of walking to a docking station. Further, we conduct a sensitivity analysis at several walking time thresholds. In terms of spatial equity, we compare accessibility with and without the equity stations to assess the effect of the initiative, before reaggregating the data for further analysis using median total household income information from the census. We find that the addition of equity stations does increase the serviced population at every threshold examined. However, although accessibility increased for the whole population, this increase was relatively modest, especially for population in the bottom 20% of median total household income. The analysis also identifies areas under-serviced within the core service area of the system.

@@

Accessibility has been applied in both a positive and normative way to inform transportation planning [@paezMeasuringAccessibilityPositive2012], but its utility to this field has evolved over the past century and has increasingly become linked with recent interest in prioritizing local proximity and modes that are suitable for local trips like walking and cycling [@levineCenturyEvolutionAccessibility2020]. As a measure of the ease of reaching potential destinations spread spatially in a given area, accessibility is relevant to PBSPs because it can identify current inequaltiy in the provision of infrastructure, as well as guide interventions that increase access for groups that are under-serviced or address gaps in transportation options. It also addresses some of the challenges of other performance measures such as level of service within a transportation network by measuring person-based indicators and exploring differences in use between population groups [@paezMeasuringAccessibilityPositive2012].

Beyond the utility derived from using shared bikes to destinations of value, an important aspect of accessibility associated with PBSPs is the distance an individual must travel to reach a bicycle share station [@kabraBikeShareSystemsAccessibility2020; @wangNewBikeShare2019]. Since the time or distance needed to reach a bicycle share station decreases the potential of accessing the program and ultimately their use of the program, the location and size (e.g., number of bicycle racks) of stations matters. Indeed, distance to bicycle share stations is associated with use [@fullerUseNewPublic2011; @wangNewBikeShare2019] and can be a barrier to using PBSPs [@fishmanBarriersBikesharingAnalysis2014]. Kabra et al. [-@kabraBikeShareSystemsAccessibility2020] found that the majority of bike share usage in Paris comes from areas within 300 m of stations, which amounts to 2-4 minutes walking by an able-bodied adult. Furthermore, similar to other public amenities affected by crowding, like health care services [e.g., @pereiraGeographicAccessCOVID192021], the utility of stations is limited by the number of bicycles that they can hold. The program may not necessarily be improved if stations are easy to reach but offer only a small number of bicycles. Likewise, more people may not opt to use the program if the supply of bicycles available at nearby stations is insufficient to meet demand. The location and size of stations is important to increase the utility of this public transportation option for more people, thus achieving vertical equity. Accessibility analyses for PBSPs constitute a positive and evaluation-based approach that also has the potential to inform equity efforts. For instance, Wang and Lindsey [-@wangNewBikeShare2019] investigated whether new or relocated bicycle share stations increased accessibility and use, which offered important insights to improve the performance of the program.

@@

In contrast, floating catchment area (FCA) methods incorporate information on capacity and demand in calculating accessibility. FCA measures have been widely employed in healthcare accessibility research and are typically calculated across two steps. In the first, a ratio of supply to demand at service locations is calculated, such as the number of beds at a hospital divided by the number of people within the catchment area of the hospital, weighted by the distance involved in reaching the facility. Next, these service level ratios are allocated back to the population centres and summarized as a measure of congested accessibility. Thus, this model does a good job of considering potential crowding or competition for services. While there have been many methodological innovations in FCA methods [for example, @delamaterSpatialAccessibilitySuboptimally2013; @luoMeasuresSpatialAccessibility2003; @luoEnhancedTwostepFloating2009; @radkeSpatialDecompositionsModeling2000; @wanThreestepFloatingCatchment2012], a recent improvement to this approach was achieved through a simple and intuitive balancing of the impedance that addressed the effects of demand and service inflation found in earlier FCA approaches [see @paezDemandLevelService2019].

When measuring accessibility, researchers have taken different approaches when it comes to aggregation of data, either by using the individual or household as the smallest unit of analysis or larger spatial zones. Previous research on bike share equity has typically used a meso- or macro-level approach with aggregate data from entire neighbourhoods or census tracts [@babagoliExploringHealthSpatial2019; @mooneyFreedomStationSpatial2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], although there are recent exceptions [@chenExploringEquityPerformance2019; @chenUnobservedHeterogeneityTransportation2021]. This is also true for studies examining correlates of bike share demand [@wangNewBikeShare2019]. Handy and Niemeier [-@handyMeasuringAccessibilityExploration1997] note that using disaggregated data in accessibility analyses provides a more accurate estimate for individuals, which is useful for addressing vertical inequity in PBSP usage.

Using the balanced floating catchment area method [@paezDemandLevelService2019], a novel approach that has not been used yet in cycling research, with disaggregate population-level data, in this paper we examine accessibility to a public bicycle share program in Hamilton, Ontario. We (1) conduct a sensitivity analysis by measuring accessibility and level of service to bike share stations at different walking time thresholds to reach a station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes; (2) explore the contribution of specific stations that were added to Hamilton's PBSP to reducing both horizontal and vertical inequities; and (3) examine whether disparities in accessibility exist according to median total household income of dissemination areas within the core service area.

@@

In fact, since the proportional allocation procedure means that any proportion of the population allocated to a station is never allocated to other stations, and conversely any level of service allocated to a population is never re-allocated elsewhere, this property is replicated for any level of aggregation. For this paper, we employ a hybrid location-based and person-based approach to calculating accessibility using disaggregate population data. In their review of accessibility measures, Geurs and van Wee [-@geursAccessibilityEvaluationLanduse2004] highlight the need for greater inclusion of individual spatio-temporal constraints but acknowledge the challenges of acquiring and analyzing person-based data. This comes after Kwan's [-@kwanSpaceTimeIntegral1998] work to show that space-time measures are  better able of capturing interpersonal differences, especially the effect of space-time constraints on individual behaviour, and are more helpful for unraveling gender/ethnic differences. Applying the balanced floating catchment area approach allows us to examine accessibility by stratifying according to median total household income, which would constitute the individual component of the accessibility measure [@geursAccessibilityEvaluationLanduse2004]. However, conducting a further sensitivity analysis to measure accessibility at different distance thresholds would help to consider potential spatio-temporal constraints. Different people may be willing to travel different distances to access a public bicycle share station.

@@

```{r interpolated-population, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Interpolated population in the Hamilton Bike Share core service area (outlined in black) and within 30 minutes of walking to the core service area (outlined in blue). Each population cell is 50-by-50 m in size. The downtown area is outlined in dark green."}
ggplot() + 
  geom_sf(data = cropped_da,
          fill = "gray",
          color = NA) + 
  #geom_sf(data = population_50x50,
  #        aes(color = population)) +
   #scale_fill_gradient(colours = "YlOrRd") +
  geom_tile(data = cbind(st_drop_geometry(population_50x50), 
                         st_coordinates(population_50x50)),
            aes(x = X, 
                y = Y, 
                fill = population)) +
  scale_fill_distiller(palette = "YlOrRd", 
                        direction = 1) +
  geom_sf(data = sobi_service,
          fill = NA,
          colour = "black") +
  geom_sf(data = community_downtown %>%
            dplyr::filter(NAME == "Hamilton"),
          fill = NA,
          color = "dark green") +
  theme(panel.background = element_rect(fill = "white"))
```

@@

```{r figure-6, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 3 minutes walk (minimum threshold) compared between current system with equity stations and the original system without equity stations."}
#accessibility plot for current system with equity stations at 3 minutes walk
accessibility_3min <- accessibility %>% 
  dplyr::filter(threshold == "3 min")

ggplot() +
  geom_sf(data = population_50x50,
          color = "gray") +
  #geom_sf(data = accessibility %>%
  #          dplyr::filter(threshold == "3 min"),
  #        aes(color = accessibility)) +
  #scale_color_distiller(palette = "PuRd", 
  #                      direction = 1) +
  geom_tile(data = cbind(st_drop_geometry(accessibility_3min), st_coordinates(accessibility_3min)),
            aes(x = X, y = Y, fill = accessibility)) +
  scale_fill_distiller(palette = "PuRd", direction = 1) +
  geom_sf(data = sobi_service,
          fill = NA) +
  facet_grid(system ~ threshold)  +
  geom_sf(data = community_downtown %>%
            dplyr::filter(NAME == "Hamilton"),
          fill = NA,
          color = "dark green") +
  # geom_sf(data =  sobi_hubs %>% 
  #           dplyr::filter(hub_type == "ERI"), 
  #         size = 1.5, shape = 21, 
  #         fill = "green") +
  theme(panel.background = element_rect(fill = "white"),
        axis.title.x=element_blank(),
        axis.title.y=element_blank())
```

@@

```{r figure-7, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 5 minutes walk (average threshold) compared between current system with equity stations and the original system without equity stations."}
#accessibility plot for current system with equity stations at 5 minutes walk
accessibility_5min <- accessibility %>% 
  dplyr::filter(threshold == "5 min")

ggplot() +
  geom_sf(data = population_50x50,
          color = "gray") +
  #geom_sf(data = accessibility %>%
  #          dplyr::filter(threshold == "5 min"),
  #        aes(color = accessibility)) +
  #scale_color_distiller(palette = "PuRd", 
  #                      direction = 1) +
  geom_tile(data = cbind(st_drop_geometry(accessibility_5min), st_coordinates(accessibility_5min)),
            aes(x = X, y = Y, fill = accessibility)) +
  scale_fill_distiller(palette = "PuRd", direction = 1) +
  geom_sf(data = sobi_service,
          fill = NA) +
  facet_grid(system ~ threshold)  +
  geom_sf(data = community_downtown %>%
            dplyr::filter(NAME == "Hamilton"),
          fill = NA,
          color = "dark green") +
  # geom_sf(data =  sobi_hubs %>% 
  #           dplyr::filter(hub_type == "ERI"), 
  #         size = 1.5, shape = 21, 
  #         fill = "green") +
  theme(panel.background = element_rect(fill = "white"),
        axis.title.x=element_blank(),
        axis.title.y=element_blank())
```

@@

```{r figure-8, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 10 minutes walk (maximum threshold) compared between current system with equity stations and the original system without equity stations."}
#accessibility plot for current system with equity stations at 10 minutes walk
accessibility_10min <- accessibility %>% 
  dplyr::filter(threshold == "10 min")

ggplot() +
  geom_sf(data = population_50x50,
          color = "lightgray") +
  #geom_sf(data = accessibility %>%
  #          dplyr::filter(threshold == "10 min"),
  #        aes(color = accessibility)) +
  #scale_color_distiller(palette = "PuRd", 
  #                      direction = 1) +
  geom_tile(data = cbind(st_drop_geometry(accessibility_10min), st_coordinates(accessibility_10min)),
            aes(x = X, y = Y, fill = accessibility)) +
  scale_fill_distiller(palette = "PuRd", direction = 1) +
  geom_sf(data = sobi_service,
          fill = NA) +
  facet_grid(system ~ threshold)  +
  geom_sf(data = community_downtown %>%
            dplyr::filter(NAME == "Hamilton"),
          fill = NA,
          color = "dark green") +
  # geom_sf(data =  sobi_hubs %>% 
  #           dplyr::filter(hub_type == "ERI"), 
  #         size = 1.5, shape = 21, 
  #         fill = "green") +
  theme(panel.background = element_rect(fill = "white"),
        axis.title.x=element_blank(),
        axis.title.y=element_blank())
```

@@

```{r figure-9, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 15 minutes walk (extreme threshold) compared between current system with equity stations and the original system without equity stations."}
#accessibility plot for current system with equity stations at 15 minutes walk
accessibility_15min <- accessibility %>% 
  dplyr::filter(threshold == "15 min")

ggplot() +
  geom_sf(data = population_50x50,
          color = "lightgray") +
  #geom_sf(data = accessibility %>%
  #          dplyr::filter(threshold == "15 min"),
  #        aes(color = accessibility)) +
  #scale_color_distiller(palette = "PuRd", 
  #                      direction = 1) +
  geom_tile(data = cbind(st_drop_geometry(accessibility_15min), st_coordinates(accessibility_15min)),
            aes(x = X, y = Y, fill = accessibility)) +
  scale_fill_distiller(palette = "PuRd", direction = 1) +
  geom_sf(data = sobi_service,
          fill = NA) +

  facet_grid(system ~ threshold)  +
  geom_sf(data = community_downtown %>%
            dplyr::filter(NAME == "Hamilton"),
          fill = NA,
          color = "dark green") +
  # geom_sf(data =  sobi_hubs %>% 
  #           dplyr::filter(hub_type == "ERI"), 
  #         size = 1.5, shape = 21, 
  #         fill = "green") +
  theme(panel.background = element_rect(fill = "white"),
        axis.title.x=element_blank(),
        axis.title.y=element_blank())
```

## Accessibility by Median Total Household Income

A unique property of the balanced floating catchment area method is that data can be reaggregated while preserving the total population in the service area and the supply at each station. This avoids demand and supply inflation, and also enables us to present findings in a way that is perhaps more intuitive to interpret. Therefore, we reaggregate population and accessibility from the 50-by-50 m population cells to the dissemination area. We then can compare accessibility to median total household income statistics from the census.

Figures 1 [\ref{fig:figure-bi-map-threshold-3}], 2 [\ref{fig:figure-bi-map-threshold-5}], 3 [\ref{fig:figure-bi-map-threshold-10}], and 4 [\ref{fig:figure-bi-map-threshold-15}] depict bivariate choropleth maps with the combine the spatial distribution of accessibility and median total household income, using tertiles for the coloring scheme. In each figure the top panel is the system without the equity stations, and the bottom panel is accessibility with the equity stations. As expected, the analysis indicates that the extreme threshold of fifteen minutes is associated with the largest number of people who are within the assumed service areas of stations. We find that stations added to Hamilton's public bicycle share program did indeed expand the spatial coverage of the system. By implementing equity stations in areas with lower median total household income in Hamilton, the PBSP attained greater horizontal equity by extending the spatial distribution of bicycles across the city. This is particularly evident at the minimum and average thresholds of three and five minutes, respectively, where the equity stations fill a number of gaps in program coverage.
